---
title: "Homework Metrics 2b: Federal Reserve's interest rates and global outstanding credit. A univariate and multivariates analysis"
author: "Valentin Auplat, Nino Laffray"
date: "May 2025"
output:
  html_document:
    df_print: paged
  pdf_document: default
---
```{r echo = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)
```

```{r echo = FALSE, message = FALSE, warning = FALSE}
library(haven)
library(dplyr)
library(knitr)
library(tidyr)
library(ggplot2)
library(stargazer)
library(Synth)
library(foreign)
library(knitr)
library(readxl)
library(purrr)
library(janitor)
library(lubridate)
library(tsibble)
```

```{r, echo = FALSE}
flatten_countries <- function(file_path) {
  # Read the sheet and skip the header row
  df <- read_excel(file_path, sheet = 3, skip = 1)

  # Rename first column
  colnames(df)[1] <- "Category"
  df <- df[c(1:20),]
  # Tag levels based on pattern rules
  df <- df %>%
    mutate(
      Level0 = ifelse(grepl("Borrowers outside", Category), Category, NA),
      Level1 = ifelse(grepl("^Of which:", Category), Category, NA),
      Level2 = ifelse(grepl("^Africa and Middle East|^Emerging Asia and Pacific|^Emerging Europe|^Latin America", Category), Category, NA),
      Level3 = ifelse(is.na(Level0) & is.na(Level1), Category, NA)
    ) %>%
    fill(Level0, Level1, Level2, .direction = "down") %>%
    select(Level0, Level1, Level2, everything())  # Keep hierarchy + data
  
  return(df)
}


flatten_instruments <- function(file_path) {
  dt <- read_excel(file_path, sheet = 3, skip = 1)
  colnames(dt)[1] <- "Category"
  dt <- dt[21:32, ]  # Adjust to the actual instrument section

  dt <- dt %>%
    mutate(
      Level0 = ifelse(grepl("By instrument|Memo: Borrowers in United States", Category), Category, NA),
      Level1 = ifelse(grepl("Borrowers outside United States|Of which: emerging market and developing economies|Non-financial borrowers", Category), Category, NA),
      Level2 = ifelse(grepl("Bank loans|Debt securities issues|Of which: government|Of which: emerging market and developing economies", Category), Category, NA)) %>%
    fill(Level0, Level1, Level2, .direction = "down") %>%
    select(Level0, Level1, Level2 , everything())
dt <- dt[-c(1,10),]
  return(dt)
}
#data <- flatten_instruments("2.xlsx")
```

```{r, echo = FALSE}
merged_data_countries <- flatten_countries("2.xlsx")
for (i in 3:34){
  data <- flatten_countries(paste0(i, ".xlsx"))
  merged_data_countries <- merge(merged_data_countries, data, by = c("Level0", "Level1", "Level2", "Level3", "Category"))
}
merged_data_countries <- merged_data_countries %>% 
  clean_names() %>%
  select(-matches("(_5|_6|_7)")) %>%
  rename_with(~ sub("^x", "", .x)) %>%
  rename_with(~ sub("(_2$|_3$|_4$)", "", .x))

merged_data_instruments <- flatten_instruments("2.xlsx")
for (i in 3:34){
  data <- flatten_instruments(paste0(i, ".xlsx"))
  merged_data_instruments <- merge(merged_data_instruments, data, by =  c("Level0", "Level1", "Level2", "Category"), all.x = TRUE)
}
merged_data_instruments <- merged_data_instruments %>% 
  clean_names() %>%
  select(-matches("(_5|_6|_7)")) %>%
  rename_with(~ sub("^x", "", .x)) %>%
  rename_with(~ sub("(_2$|_3$|_4$)", "", .x))

df_long <- merged_data_countries %>%
  select(-level0, -level1, -level2, -level3) %>%
  pivot_longer(
    cols = -category, # all columns except 'category'
    names_to = "quarter",
    values_to = "value"
  ) %>%
  arrange(category, quarter)

FED <- read.csv("FEDFUNDS.csv")
FED <- FED %>%
  mutate(
    observation_date = ymd(observation_date),
    quarter = quarter(observation_date),
    year = year(observation_date),
    observation_date = paste0(year, "-q", quarter)
  ) %>%
  select(-year, -quarter)
FED <- FED %>%
  rename(value = FEDFUNDS) %>%
  rename(quarter = observation_date) %>%
  mutate(category = "FED_rate")
FED <- FED %>%
    mutate(quarter = gsub("-", "_", FED$quarter))

df_long <- rbind(df_long, FED) %>%
  filter(quarter >= "2000-q1") %>%
  filter(quarter <= "2023-q4") %>%
  pivot_wider(names_from = category, values_from = value)
```
# Question 1: Plots of the variables of interest.
```{r, echo = FALSE}
df_long <- ts(df_long, start=c(2000, 1), frequency=4)
plot(df_long[, "FED_rate"], main = "Evolution of FED's interest rate over time", xlab = "Time", ylab = "Fed's interest rate")
mtext("Source: Federal funds effective rate, Economic Data, St. Louis FED", side = 1, line = 4, adj = 1, cex = 0.8)

plot(df_long[, "Borrowers outside United States"], main = "Evolution of debt denominated in USD (borrowers outside US)", xlab = "Time", ylab = "Debt denominated in USD (M USD)")
mtext("Source: Total credit to non-bank borrowers by currency of denomination (USD) - GLI_E2_1, Bank of International Settlements", side = 1, line = 4, adj = 1, cex = 0.57)

plot(df_long[,c("Emerging Asia and Pacific", "Emerging Europe", "Latin America")], plot.type="s", col=c("red","green", "blue"), main = "Evolution of debt denominated in USD (aggregates)", xlab = "Time", ylab = "Debt denominated in USD (M USD)")
legend("topleft", legend=c("Emerging Asia and Pacific","Emerging Europe", "Latin America") ,col=c("red","green", "blue"), lty=1, bty="n")
mtext("Source: Total credit to non-bank borrowers by currency of denomination (USD) - GLI_E2_1, Bank of International Settlements", side = 1, line = 4, adj = 1, cex = 0.57)
```
Stationarity cannot be assumed from these graphs. We see that variables present trends, but not cycles. For this first exercise, we are mostly interested in the interest rate of the Federal Reserve and the amount of credit borrowed outside the United States denominated in USD.

# Question 2: Stationarity tests.
```{r, echo = FALSE}
library(urca)
library(zoo)
library(forecast)
library(kableExtra)
df_long <- ts(df_long, start=c(2000, 1), end=c(2020, 1), frequency=4)
# 2.1 ADF Test (Augmented Dickey-Fuller)
adf_test <- ur.df(df_long[,c("FED_rate")], type = "trend", lags = 2)  # Try "drift" or "none" too
summary(adf_test)
# Extract values
test_stat <- round(adf_test@teststat[1], 3)
crit_vals <- round(adf_test@cval[1, ], 3)

# Reformat into a long data frame
adf_table <- data.frame(
  Statistic = c("Test Statistic", "Critical Value (1%)", "Critical Value (5%)", "Critical Value (10%)"),
  Value = c(test_stat, crit_vals["1pct"], crit_vals["5pct"], crit_vals["10pct"])
)

# Print with clean headers
kable(adf_table, caption = "Augmented Dickey-Fuller Test Results", booktabs = TRUE, align = "lc") %>%
  kable_styling(latex_options = c("hold_position"), font_size = 10)
```
Although the visual plot of the FED rate shows distinct level shifts and prolonged flat regimes (e.g., 2008–2016), the ADF test result shows that the process is stationary once we control for trend and include appropriate lags.

This means the visual "non-stationarity" may be deterministic trend or policy-driven shifts, but not a stochastic unit root.


```{r}
# 2.1 ADF Test (Augmented Dickey-Fuller)
adf_test <- ur.df(df_long[,c("Borrowers outside United States")], type = "trend", lags = 4)  # Try "drift" or "none" too
summary(adf_test)
```
Idem.


```{r}
# 2.3 KPSS Test (Stationarity)
kpss_test <- ur.kpss(df_long[,c("FED_rate")], type = "tau", lags = "long")
summary(kpss_test)
```

```{r}
# 2.3 KPSS Test (Stationarity)
kpss_test <- ur.kpss(df_long[,c("Borrowers outside United States")], type = "tau", lags = "long")
summary(kpss_test)
```

```{r}
library(forecast)
Acf(df_long[,c("FED_rate")], lag.max=20) # the MA part could be q=3, but a lot of non-null higher order lags
Pacf(df_long[,c("FED_rate")], lag.max=20) # the AR part could be p=3 with the full sample
ar3 <- Arima(df_long[,c("FED_rate")], order=c(1,2,0), include.constant=TRUE) # start with the AR part, assume integration order 0 (stationary)
summary(ar3) # last lag and constant are significant
plot(cbind(ar3$x, ar3$fitted), plot.type="s", col=c("black","red"), lty=1)
legend("topright", legend=c("GDP (QoQ growth rate)","AR(3)") ,col=c("black","red"), lty=1, bty="n")


ma3 <- Arima(df_long[,c("FED_rate")], order=c(0,0,3), include.constant=TRUE) # continue with the MA part
summary(ma3) # last lag and the constant are significant.
plot(cbind(ma3$x,ma3$fitted), plot.type="s", col=c("black","red"), lty=1)
legend("topright", legend=c("GDP (QoQ growth rate)","MA(3)") ,col=c("black","red"), lty=1, bty="n")


# Fit AR models
ar1 <- Arima(df_long[,c("FED_rate")], order = c(1, 0, 0), include.constant = TRUE)
ar2 <- Arima(df_long[,c("FED_rate")], order = c(2, 0, 0), include.constant = TRUE)
ar3 <- Arima(df_long[,c("FED_rate")], order = c(3, 0, 0), include.constant = TRUE)

# Create comparison table
model_comparison <- data.frame(
  Model = c("AR(1)", "AR(2)", "AR(3)"),
  AIC = c(AIC(ar1), AIC(ar2), AIC(ar3)),
  BIC = c(BIC(ar1), BIC(ar2), BIC(ar3)),
  LogLikelihood = c(logLik(ar1), logLik(ar2), logLik(ar3))
)

print(model_comparison)
```